---
title: "Hospital Capacity Planning Using Discrete Event Simulation: Simmer" 
author:
  - name: Thomas Bartz-Beielstein
    affiliation: IDEA, TH Köln
    address:
    - line 1
    - line 2
    email:  author1@work
  - name: Frederik Rehbach
    affiliation: IDEA, TH Köln
    address:
    - line 1
    - line 2
    email:  author2@work
  - name: Olaf Mersmann
    affiliation: Affiliation
    address:
    - line 1
    - line 2
    email:  author2@work
  - name: Eva Bartz
    affiliation: Affiliation
    address:
    - line 1
    - line 2
    email:  author2@work
abstract: >
  An abstract of less than 150 words.
preamble: |
  % Any extra LaTeX you need in the preamble
header-includes:
   - \usepackage{bbm}
   - \usepackage{glossaries}
   - \usepackage{xspace}
   - \newacronym{crn}{CRN}{Common Random Numbers}
   - \newcommand{\healed}{\texttt{Healed}\xspace}
   - \newcommand{\bed}{\texttt{bed}\xspace}
   - \newcommand{\nurse}{\texttt{nurse}\xspace}
   - \newcommand{\sick}{\texttt{sick}\xspace}
   - \newcommand{\amntDaysSickness}{\texttt{amntDaysSickness}\xspace}
   - \newcommand{\factorVentilation}{\texttt{factorVentilation}\xspace}
   - \newcommand{\factorIntensive}{\texttt{factorIntensive}\xspace}
   - \newcommand{\factorHospital}{\texttt{factorHospital}\xspace}
   - \newcommand{\intensiveBed}{\texttt{intensiveBed}\xspace}
   - \newcommand{\intensiveBedVentilation}{\texttt{intensiveBedVentilation}\xspace}
   - \newcommand{\FactorPatientsVentilationToDeath}{\texttt{FactorPatientsVentilationToDeath}\xspace}
   - \newcommand{\FactorPatientsIntensiveToVentilation}{\texttt{FactorPatientsIntensiveToVentilation}\xspace}
   - \newcommand{\FactorPatientsIntensiveToDeath}{\texttt{FactorPatientsIntensiveToDeath}\xspace}
   - \newcommand{\FactorPatientsIntensiveToAftercare}{\texttt{FactorPatientsIntensiveToAftercare}\xspace}
   - \newcommand{\FactorPatientsNormalToIntensive}{\texttt{FactorPatientsNormalToIntensive}\xspace} 
   - \newcommand{\FactorPatientsNormalToDeath}{\texttt{FactorPatientsNormalToDeath}\xspace}
   - \newcommand{\FactorPatientsNormalToVentilation}{\texttt{FactorPatientsNormalToVentilation}\xspace}
   - \newcommand{\FactorPatientsNormalToHealed}{\texttt{FactorPatientsNormalToHealed}\xspace}
   - \newcommand{\FactorPatientsHospitalToNormal}{\texttt{FactorPatientsHospitalToNormal}\xspace}  
   - \newcommand{\FactorPatientsHospitalToIntensive}{\texttt{FactorPatientsHospitalToIntensive}\xspace}   
   - \newcommand{\FactorPatientsHospitalToVentilation}{\texttt{FactorPatientsHospitalToVentilation}\xspace}  
   - \newcommand{\FactorPatientsHospitalToDeath}{\texttt{FactorPatientsHospitalToDeath}\xspace}  
   - \newcommand{\FactorPatientsInfectedToAmbulant}{\texttt{FactorPatientsInfectedToAmbulant}\xspace} 
   - \newcommand{\FactorPatientsInfectedToHealthy}{\texttt{FactorPatientsInfectedToHealthy}\xspace} 
   - \newcommand{\FactorPatientsInfectedToHospital}{\texttt{FactorPatientsInfectedToHospital}\xspace} 
   - \newcommand{\AmntNursesPerPatient}{\texttt{AmntNursesPerPatient}\xspace}
   - \newcommand{\AmntInfectedPerDay}{\texttt{AmntInfectedPerDay}\xspace}
   - \newcommand{\AmntInfectedPerDayCum}{\texttt{AmntInfectedPerDayCum}\xspace}
   - \newcommand{\AmntDaysVentilationToAftercare}{\texttt{AmntDaysVentilationToAftercare}\xspace}
   - \newcommand{\AmntDaysIntensiveToAftercare}{\texttt{AmntDaysIntensiveToAftercare}\xspace}
   - \newcommand{\AmntDaysIntensiveToDeath}{\texttt{AmntDaysIntensiveToDeath}\xspace}
   - \newcommand{\AmntDaysIntensiveToVentilation}{\texttt{AmntDaysIntensiveToVentilation}\xspace}
   - \newcommand{\AmntDaysAmbulant}{\texttt{AmntDaysAmbulant}\xspace}
   - \newcommand{\StartDate}{\texttt{StartDate}\xspace}
   - \newcommand{\aftercare}{\texttt{aftercare}\xspace}
   - \newcommand{\intensive}{\texttt{intensive}\xspace}
   - \newcommand{\ventilation}{\texttt{ventilation}\xspace}
   - \newcommand{\ambulant}{\texttt{ambulant}\xspace}
   - \newcommand{\normalStation}{\texttt{normalStation}\xspace}
   - \newcommand{\hospital}{\texttt{hospital}\xspace}
   - \newcommand{\infected}{\texttt{infected}\xspace}
   - \newcommand{\death}{\texttt{death}\xspace}
   - \newcommand{\transferout}{\texttt{transferout}\xspace} 
   - \newcommand{\intensiveAfter}{\texttt{intensiveAfter}\xspace}
date: "`r Sys.Date()`"
output:
   rmarkdown::html_vignette:
      fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Hospital Capacity Planning Using Discrete Event Simulation: Simmer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---   



## The `simmer` Model

### Warmup Period (Burn in)

* The babsim model does not consider occupied beds. 
* It does not simulate a warmup period.

### Elements of the Hospital System

* TODO: Describe the patient trajectory.
* Factors and durations, e.g., 
`FactorPatientsInfectedToHealthy = 0.831` and `AmntDaysInfectedToHealed = 20.5`.
* TODO: Create a table to list states and transition probabilities and durations.

* The states, durations, and transition probabilities are visualized in Fig.~\ref{fig:20200803states}

```{r simmerStates, echo=FALSE, fig.cap='Simmer States.', out.width='5in', fig.align='center', fig.pos='htbp'}
knitr::include_graphics('20200803states.pdf')
```

## Performing simmer Simulations

```{r, include = FALSE, eval = TRUE}
simRepeats <- 25
```
We will perform  `r simRepeats` simulation runs (repeats with different seeds).

* simmer counters

The following counter are used:
\begin{itemize}
\item "No Hospital Required"
\item "Dead"
\item \healed
\end{itemize}



### simmer Activities


```{r, include= FALSE}
 minVal <- 1e-6 
  # FactorPatientsInfectedToHealthy = 0.831, # wird nicht genutzt, sondern 1 - x17, ok
    # AmntDaysInfectedToHealed = 20.5, # wird nicht genutzt, ok
    #
    # 2 Infiziert -> Krankenhaus Infected -> Hospital
    FactorPatientsInfectedToHospital = 0.169 # x17 
    AmntDaysInfectedToHospital = 8.4 # x1
    #
    # Not used: Infected -> Ambulant
    # FactorPatientsInfectedToAmbulant = minVal
    # AmntDaysInfectedToAmbulant = minVal
    #
    # 3 Krankenhaus -> Normalstation Hospital -> Normal
    FactorPatientsHospitalToNormal = minVal # fehlt (Differenz, ok)
    AmntDaysHospitalToNormal = minVal # nicht genutzt
    #
    # 4 Krankenhaus -> Intensiv ohne Beatmung Hospital -> Intensive
    FactorPatientsHospitalToIntensive = 0.04  #0.012, # x18
    AmntDaysHospitalToIntensiv = minVal # nicht genutzt
    #
    # 5 Krankenhaus -> Intensiv mit Beatmung Hospital -> Ventilation
    FactorPatientsHospitalToVentilation = 0.036 #x19
    AmntDaysHospitalToVentilation = minVal # nicht genutzt
    #
    #6 Normalstation -> Healthy
    FactorPatientsNormalToHealthy = minVal # fehlt (Differenz, ok)
    AmntDaysNormalToHealthy = 11.6 #x2
    #
    #7 Normalstation ->  Intensive
    FactorPatientsNormalToIntensive = 0.0506 #x20 
    AmntDaysNormalToIntensive = 1.25 # x3
    #
    #8 Normalstation ->  Ventilation
    FactorPatientsNormalToVentilation = 0.1013 #x21
    AmntDaysNormalToVentilation = 3.63 #x4
    #
    #9 Normalstation -> Verstorben Normal -> Death
    FactorPatientsNormalToDeath = 0.139 # x22
    AmntDaysNormalToDeath  = 11.4 # x5
    #
    #10 Intensiv ohne Beatmung -> Aftercare
    FactorPatientsIntensiveToAftercare = 0.25 # dependent, not used
    AmntDaysIntensiveToAftercare = 7.0 # x6
    #
    #11 Intensiv ohne Beatmung  -> Ventilation
    FactorPatientsIntensiveToVentilation = 0.25 # x23
    AmntDaysIntensiveToVentilation = 2.0 # x7
    #
    #12 Intensiv ohne Beatmung -> Death
    FactorPatientsIntensiveToDeath = 0.25 # x24
    AmntDaysIntensiveToDeath = 2.0 # x8
    #
    #12a Intensiv ohne Beatmung ->  Healthy
    FactorPatientsIntensiveToHealthy = 0.25 # x25
    AmntDaysIntensiveToHealthy = 13.0 # x9
    #
    #13 Intensiv mit Beatmung -> Nachbehandlung Ventilation -> Aftercare
    FactorPatientsVentilationToAftercare = 0.08 # nicht benoetigt, Differenz, ok
    AmntDaysVentilationToAftercare = 9 # erhoeht, da mit #14 zusammen # x10
    #
    #14 Intensiv mit Beatmung ->  IntensiveAfter
    FactorPatientsVentilationToIntensiveAfter = 0.42 # x26
    AmntDaysVentilationToIntensiveAfter = 23.0 #  x11
    #
    #15 Intensiv mit Beatmung -> Death
    FactorPatientsVentilationToDeath = 0.50  # x27
    AmntDaysVentilationToDeath = 16 # x12
    #
    #16 Nachbehandlung -> Healthy
    FactorPatientsAftercareToHealthy = 1 # 100% not used
    AmntDaysAftercareToHealthy = 21 # x30
    #
    #17 I Intensiv ohne Beatmung II: IntensiveAfter -> Aftercare
    FactorPatientsIntensiveAfterToAftercare = 0.50 # dependent, not used
    AmntDaysIntensiveAfterToAftercare = 7.0 # x13
    
    #17 II Intensiv ohne Beatmung II: IntensiveAfter -> Healthy
    FactorPatientsIntensiveAfterToHealthy = 0.50 # x28
    AmntDaysIntensiveAfterToHealthy = 18.0 # x14
    
    #18 Intensiv ohne Beatmung II: IntensiveAfter -> Death
    FactorPatientsIntensiveAfterToDeath = 1e-5 # x29
    AmntDaysIntensiveAfterToDeath = 1 # x15
    #
    GammaShapeParameter = 1.0 # x16
    # Not used 
    AmntDaysAmbulant = minVal # small value, if 0
    # is in conf: maxCapacity = 1e4
    AmntNursesPerPatient = 1
    # Risks based on age
    RiskFactorA = 0.02048948 #x31
    RiskFactorB = 0.01 #32
    RiskMale = 2.0 #x33
```


Durations are defined as activities, i.e., completion times are random draws from probability distributions.

* Assuming exponentially distributed durations, then GammaShapeParameter should be 1.

### Simmer Trajectories

* Trajectory `transferout`

Patients that do not require hospital are sent home.
Here, we model state changes from the state infected.

*  Trajectory `Dead`

Patients that die.

* Trajectory `aftercare`

An aftercare patient requires one bed.
The patient stays on average `r AmntDaysAftercareToHealthy` days in this bed.
Afterwards, the counter \healed is incremented.

* Trajectory `intensiveAfter`

* Trajectory `ventilation`

The resource \bed is released and the resource \intensiveBedVentilation is seized.
The following trajectories can be reached:
\begin{enumerate}
\item 
Patients move to trajectory \death with probability \FactorPatientsVentilationToDeath  after 
 days.
\item 
Patient move to trajectory \aftercare with probability (1- \FactorPatientsVentilationToDeath) 
after \AmntDaysVentilationToAftercare days.
\end{enumerate}

* Trajectory `intensive`

The resource \bed is released and one resource \intensiveBed is seized.
The following trajectories can be reached:
\begin{enumerate}
\item Patients move from \intensive to trajectory \ventilation with probability \FactorPatientsIntensiveToVentilation after \AmntDaysIntensiveToVentilation days
\item Patients move from \intensive to trajectory \death with probability \FactorPatientsIntensiveToDeath  after \AmntDaysIntensiveToDeath days
\item Patients move from \intensive to trajectory \aftercare with probability \FactorPatientsIntensiveToAftercare  after \AmntDaysIntensiveToAftercare days
\end{enumerate}


* Trajectory `ambulant`

The resource \bed is seized.
The resource \nurse is seized \AmntNursesPerPatient times.
The following trajectories can be reached:
\begin{enumerate}
\item Patients move from \ambulant to \healed after \AmntDaysAmbulant days
\end{enumerate}

* Trajectory `normalStation`

Normal Bed in the Hospital.
The resource \bed is seized.
The resource \nurse is seized `r AmntNursesPerPatient` times.
The following trajectories can be reached:
\begin{enumerate}
\item Patients move from \normalStation to trajectory \intensive with probability \FactorPatientsNormalToIntensive = `r FactorPatientsNormalToIntensive` after `r AmntDaysIntensiveToVentilation` days
\item Patients move from \normalStation to trajectory \death with probability \FactorPatientsNormalToDeath = `r FactorPatientsNormalToDeath` after `r AmntDaysNormalToDeath` days
\item Patients move from \normalStation to trajectory \ventilation with probability \FactorPatientsNormalToVentilation = `r FactorPatientsNormalToVentilation` after `r AmntDaysNormalToVentilation` days
\item Patients move from \normalStation to \healed with probability \FactorPatientsNormalToHealed = `r (1- FactorPatientsNormalToIntensive - FactorPatientsNormalToDeath - FactorPatientsNormalToVentilation) ` after `r AmntDaysNormalToHealthy` days
\end{enumerate}

* Trajectory `hospital`

Infected patient that comes to the hospital is immediately sent to a specific station.
No resources are seized in this state.
The following trajectories can be reached:
\begin{enumerate}
\item Patients move from \hospital to trajectory \intensive with probability \FactorPatientsHospitalToIntensive = `r FactorPatientsHospitalToIntensive`
\item Patients move from \hospital to trajectory \ventilation with probability \FactorPatientsNormalToVentilation = `r FactorPatientsHospitalToVentilation` 
\item Patients move from \hospital to trajectory \normalStation with probability \FactorPatientsHospitalToNormal = `r 1 -(FactorPatientsHospitalToIntensive + FactorPatientsHospitalToVentilation)`
\end{enumerate}


* Trajectory `infected`

Initial Trajectory.
Infected person becomes sick after `r AmntDaysInfectedToHospital` days and goes to the \hospital.
No resources are seized in this state.
The following trajectories can be reached:
\begin{enumerate}
\item Patients move from \infected to trajectory \transferout with probability \FactorPatientsInfectedToHealthy = `r 1- FactorPatientsInfectedToHospital` 
\item Patients move from \infected to trajectory \hospital with probability \FactorPatientsInfectedToHospital = `r FactorPatientsInfectedToHospital`
\end{enumerate}

### Starting One Simulation

* The simmer babsim simulations start with arrivalTimes that are based on the number of \infected persons (on a daily basis), not \sick persons! 

```{r, eval = FALSE, include= FALSE}
simmer("Simulation") %>%
  add_dataframe("newInfected", infected, arrivalTimes, time = "absolute") %>%
  add_resource("bed", Amnt_Normal_Beds) %>%
  add_resource("intensiveBed", Amnt_Intensive_Care_Beds) %>%
  add_resource("intensiveBedVentilation",
               Amnt_Intensive_Care_Beds_Ventilation) %>%
  add_resource("nurse", Amnt_Nurses) %>%
  #add_resource("doctor",Amnt_Doctors) %>%
  run()
```

*  Simulate n replicates

```{r, eval = FALSE, include= FALSE}
envs <- lapply(1:simRepeats, function(i) {
  simmer("Simulation") %>%
    add_dataframe("newInfected", infected, arrivalTimes, time = "absolute") %>%
    add_resource("bed", Amnt_Normal_Beds) %>%
    add_resource("intensiveBed", Amnt_Intensive_Care_Beds) %>%
    add_resource("intensiveBedVentilation",
                 Amnt_Intensive_Care_Beds_Ventilation) %>%
    add_resource("nurse", Amnt_Nurses) %>%
    #add_resource("doctor",Amnt_Doctors) %>%
    run()
})
```

*  Results
```{r, eval = FALSE, include= FALSE}
AmntInfected = list( infectedPerDayCum, infectedPerDay)
```

Extract resources from simulation

```{r, eval = FALSE, include= FALSE}
resources <- get_mon_resources(envs)
## simple resource plot
print(plot(resources, steps=TRUE))
```

* Summary Statistics

```{r, eval = FALSE, include= FALSE}
resources$time <- round(resources$time) ## Round Time to single days!

## If there are replicates, look for median, worst and best case scenario
resourcesMaxSystem <- resources %>% group_by(resource, time) %>%  mutate(upper = max(system)) %>%
    mutate(lower = min(system)) %>% mutate(med = median(system))

## Cutoff date for plotting
resourcesMaxSystem <- resourcesMaxSystem %>% filter(time <= (observedPeriod + 
                                                        as.numeric(as.Date(StartDateSimulation)- as.Date(StartDate))))

resourcesMaxSystem$date<-as.POSIXct((resourcesMaxSystem$time)*24*60*60,origin=StartDate)

```



```{r, eval = FALSE, include= FALSE}
## Add simulations from other sources to plots
resourcesMaxSystem$source = "BuB"
resourcesMaxSystem <- resourcesMaxSystem %>% filter(resource != "nurse")

resourcesMaxSystem <- bind_rows(resourcesMaxSystem,
                                getDataFromFixedNumbers(0.027,0,0.013,  sickPerDay,"Köln"))
resourcesMaxSystem <- bind_rows(resourcesMaxSystem,
                                getDataFromFixedNumbers(0.14,0,0.03,  sickPerDay,"WHO"))
resourcesMaxSystem <- bind_rows(resourcesMaxSystem,
                                getDataFromFixedNumbers(0.1347,0.004,0.0171,  sickPerDay,"OBK"))
```


```{r, eval = FALSE, include= FALSE}
{p <- ggplot(resourcesMaxSystem,aes(x=date,y=med, color = source)) +
          geom_line() + 
          geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.3) + 
          facet_wrap(facets = vars(resource)) +
          labs(x="Date",y="Usage")
print(p)}
```

