---
title: "Hospital Capacity Planning Using Discrete Event Simulation: OBK" 
author:
  - name: Thomas Bartz-Beielstein
    affiliation: IDEA, TH Köln
    address:
    - line 1
    - line 2
    email:  author1@work
  - name: Frederik Rehbach
    affiliation: IDEA, TH Köln
    address:
    - line 1
    - line 2
    email:  author2@work
  - name: Olaf Mersmann
    affiliation: Affiliation
    address:
    - line 1
    - line 2
    email:  author2@work
  - name: Eva Bartz
    affiliation: Affiliation
    address:
    - line 1
    - line 2
    email:  author2@work
abstract: >
  An abstract of less than 150 words.
preamble: |
  % Any extra LaTeX you need in the preamble
header-includes:
   - \usepackage{bbm}
   - \usepackage{glossaries}
   - \usepackage{xspace}
   - \newacronym{crn}{CRN}{Common Random Numbers}
   - \newcommand{\healed}{\texttt{Healed}\xspace}
   - \newcommand{\bed}{\texttt{bed}\xspace}
   - \newcommand{\nurse}{\texttt{nurse}\xspace}
   - \newcommand{\sick}{\texttt{sick}\xspace}
   - \newcommand{\amntDaysSickness}{\texttt{amntDaysSickness}\xspace}
   - \newcommand{\factorVentilation}{\texttt{factorVentilation}\xspace}
   - \newcommand{\factorIntensive}{\texttt{factorIntensive}\xspace}
   - \newcommand{\factorHospital}{\texttt{factorHospital}\xspace}
   - \newcommand{\intensiveBed}{\texttt{intensiveBed}\xspace}
   - \newcommand{\intensiveBedVentilation}{\texttt{intensiveBedVentilation}\xspace}
   - \newcommand{\FactorPatientsVentilationToDeath}{\texttt{FactorPatientsVentilationToDeath}\xspace}
   - \newcommand{\FactorPatientsIntensiveToVentilation}{\texttt{FactorPatientsIntensiveToVentilation}\xspace}
   - \newcommand{\FactorPatientsIntensiveToDeath}{\texttt{FactorPatientsIntensiveToDeath}\xspace}
   - \newcommand{\FactorPatientsIntensiveToAftercare}{\texttt{FactorPatientsIntensiveToAftercare}\xspace}
   - \newcommand{\FactorPatientsNormalToIntensive}{\texttt{FactorPatientsNormalToIntensive}\xspace} 
   - \newcommand{\FactorPatientsNormalToDeath}{\texttt{FactorPatientsNormalToDeath}\xspace}
   - \newcommand{\FactorPatientsNormalToVentilation}{\texttt{FactorPatientsNormalToVentilation}\xspace}
   - \newcommand{\FactorPatientsNormalToHealed}{\texttt{FactorPatientsNormalToHealed}\xspace}
   - \newcommand{\FactorPatientsHospitalToNormal}{\texttt{FactorPatientsHospitalToNormal}\xspace}  
   - \newcommand{\FactorPatientsHospitalToIntensive}{\texttt{FactorPatientsHospitalToIntensive}\xspace}   
   - \newcommand{\FactorPatientsHospitalToVentilation}{\texttt{FactorPatientsHospitalToVentilation}\xspace}  
   - \newcommand{\FactorPatientsHospitalToDeath}{\texttt{FactorPatientsHospitalToDeath}\xspace}  
   - \newcommand{\FactorPatientsInfectedToAmbulant}{\texttt{FactorPatientsInfectedToAmbulant}\xspace} 
   - \newcommand{\FactorPatientsInfectedToHealthy}{\texttt{FactorPatientsInfectedToHealthy}\xspace} 
   - \newcommand{\FactorPatientsInfectedToHospital}{\texttt{FactorPatientsInfectedToHospital}\xspace} 
   - \newcommand{\AmntNursesPerPatient}{\texttt{AmntNursesPerPatient}\xspace}
   - \newcommand{\AmntInfectedPerDay}{\texttt{AmntInfectedPerDay}\xspace}
   - \newcommand{\AmntInfectedPerDayCum}{\texttt{AmntInfectedPerDayCum}\xspace}
   - \newcommand{\AmntDaysVentilationToAftercare}{\texttt{AmntDaysVentilationToAftercare}\xspace}
   - \newcommand{\AmntDaysIntensiveToAftercare}{\texttt{AmntDaysIntensiveToAftercare}\xspace}
   - \newcommand{\AmntDaysIntensiveToDeath}{\texttt{AmntDaysIntensiveToDeath}\xspace}
   - \newcommand{\AmntDaysIntensiveToVentilation}{\texttt{AmntDaysIntensiveToVentilation}\xspace}
   - \newcommand{\AmntDaysAmbulant}{\texttt{AmntDaysAmbulant}\xspace}
   - \newcommand{\StartDate}{\texttt{StartDate}\xspace}
   - \newcommand{\aftercare}{\texttt{aftercare}\xspace}
   - \newcommand{\intensive}{\texttt{intensive}\xspace}
   - \newcommand{\ventilation}{\texttt{ventilation}\xspace}
   - \newcommand{\ambulant}{\texttt{ambulant}\xspace}
   - \newcommand{\normalStation}{\texttt{normalStation}\xspace}
   - \newcommand{\hospital}{\texttt{hospital}\xspace}
   - \newcommand{\infected}{\texttt{infected}\xspace}
   - \newcommand{\death}{\texttt{death}\xspace}
   - \newcommand{\transferout}{\texttt{transferout}\xspace} 
   - \newcommand{\intensiveAfter}{\texttt{intensiveAfter}\xspace}
date: "`r Sys.Date()`"
output:
   rmarkdown::html_vignette:
      fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Hospital Capacity Planning Using Discrete Event Simulation: OBK}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---   


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

* Resource and capacity planning for hospitals.
* Paper from Lawton. \cite{Desv20a}


## Materials and Methods

```{r loadBabsim, include=FALSE}
require("vctrs")
require("slider")
require("dplyr")
require("parallel")
require("simmer")
require("simmer.plot")
require("ggplot2")
require("lubridate")
require("tidyverse")
require("plotly")
require("babsim.hospital")
```


```{r SPOT, eval = FALSE}
## install.packages("devtools")
## devtools::install_github("r-lib/devtools")
url <- "http://owos.gm.fh-koeln.de:8055/bartz/spot.git"
devtools::install_git(url = url)

library(SPOT)
packageVersion("SPOT")
```


# OBK Data

## Data

* simData contains information about the simulation period, i.e., the time
interval used by the babsim simulator.
* The correspondig dates are stored in the configuration as 
    * simulationDates: StartDate and End Date
* fieldData contains information about the period when information about the
  ground truth is available, e.g., the real number of bed, intensiveBed,
  intensiveBedVentilation resources.
* `dataCovidBeds20200624` is a data frame of COVID-19 cases with 114 obs. of  
6 variables (114 rows and 6 columns), which was manually compiled from the OBK Excel files:
   * bed (int)
   * intensiveBed (int) 
   * intensiveBedVentilation (int)
   * Day (Date, format: "2020-03-03" "2020-03-04" "2020-03-05" "2020-03-06" ...)
   * Infected (num) 
   * Sick (num)

## Set up simData

The function `getObkData` generates a list that contains two data frames: 
1. simData and 
2. field data.
```{r, getObkData}
data <- getObkData()
simData <- data$simData
fieldData  <- data$fieldData
conf <- babsimToolsConf()
```

Perform the simulation

```{r, performObKSim}
conf <- getConfFromData(conf = conf,
                        simData = simData,
                        fieldData = fieldData)
set.seed(conf$seed)
para <- babsimHospitalPara()
arrivalTimes <- getArrivalTimes(simData$Infected)
envs <- babsimHospital(arrivalTimes = arrivalTimes,
                    conf = conf,
                    para = para
                    )
```

Plot resource usage

```{r, plotMonResources}
resources <- get_mon_resources(envs)
resources$capacity <- resources$capacity/1e5
plot(resources, metric = "usage", conf$ResourceNames, items = "server")
```

Get real data (field data)

```{r, CombineSimField}
fieldEvents <- getRealBeds(data = fieldData,
                           resource=c('bed', 'intensiveBed', 'intensiveBedVentilation'))
res <- getDailyMaxResults(conf=conf, envs = envs,  fieldEvents = fieldEvents)
## save as default:
(errDefault <- getError(conf=conf, res))
```


```{r, plotSimField}
plotDailyMaxResults(res)
```

## Optimization (to be used from the command line in the folder Run.d)
```{r, eval = FALSE}
expName <- "obkNoise35"
require("SPOT")
require("babsim.hospital")
require("simmer")
##########################################################
data <- getObkData()
###########################################################
## n = number of function evaluations, Anzahl der 30-dim x Vektoren, die getestet werden
n = 35
## k = Groesse des initialen Designs für SPOT (k < n)
k = 30
## m = Anzahl der Simulationswiederholungen für Simnf
m = 2
## parallele Auswertung?
PARALLEL = FALSE
##
ICU = FALSE
##
PERCORES = 0.5
## Anzahl der Punkte, die von SPOT fuer das Metamodell verwendet werden. Sollte
## fuer Kriging nicht zu gross gewählt werden (<100):
SUBSET = 50
#############################################################
x0 <- getStartParameter()
bounds <- getBounds()
conf = babsimToolsConf()
set.seed(conf$seed)
conf$parallel = PARALLEL
conf$simRepeats = m
conf$ICU = ICU
conf$percCores = PERCORES
conf$logLevel =1
FILENAME <-  paste0("../Results.d/run", expName, ".RData")
assign( expName, 
        spot(
          x = x0,
          fun = funWrapOptimizeSim,
          lower = bounds$lower,
          upper = bounds$upper,
          control = list(
            funEvals = n,
            noise = TRUE,
            optimizer = optimLBFGSB,
            model =  buildKriging,
            plots = FALSE,
            progress = TRUE,
            subsetSelect = selectN,
            subsetControl = list(N = SUBSET),
            designControl = list(size = k)
          ),
          conf = conf,
          data = data
        )
)
# save(list = eval(parse(text = "expName")) , file = FILENAME)
```



### The OBK Data (Real-world Data)


* OBK data are included in the `babsim` package, which can be loaded as follows:

```{r}
amntDaysSickness <- 20
obk <- dataCovidBeds20200624
obk$Sick2 <-  slide_dbl(obk$Infected, ~sum(.x), .before = (amntDaysSickness -1))
str(obk)
sum(obk$Sick- obk$Sick2)
```


```{r}
obk$InfCum <- cumsum(obk$Infected)
```


```{r obkData, fig.show='hold', fig.cap="OBK data"}
{plot(obk$Day, obk$InfCum, type="l", log="y", ylim=c(1,500))
lines(obk$Day, obk$Infected + 1e-6)}
```
```{r}
obk$InfectedSlideWeek <-  slide_dbl(obk$Infected, ~mean(.x), .before = (7 -1))
obk$InfectedSlide2Weeks <-  slide_dbl(obk$Infected, ~mean(.x), .before = (14 -1))
```

```{r}
{plot(obk$Day, obk$Infected, type="l")
lines(obk$Day, obk$InfectedSlideWeek, col = "green")
lines(obk$Day, obk$InfectedSlide2Weeks, col = "grey")
legend('topright', lty = 1, c("infected", "infectedWeek",  "infected2Weeks"), col = c("black", "green",  "grey"))
}
```


```{r}
{plot(obk$Day, obk$Infected, type="l")
lines(obk$Day, obk$InfectedSlideWeek, col = "green")
lines(obk$Day, obk$bed, col = "blue")
lines(obk$Day, obk$intensiveBed, col = "orange")
lines(obk$Day, obk$intensiveBedVentilation, col = "red")
legend('topright', lty = 1, c("infected", "infectedWeek",  "normal", "intens", "vent"), col = c("black", "green",  "blue", "orange", "red"))
}
```


### Performing Simulations Using Real Data

*  Simulation with Real Data

To define the simulation period, the `obk` data will be used:
\begin{itemize}
\item `StartDate`: the day, logging of real data starts, e.g., "2020-03-03"
\item `EndDate`: last day for simulations, e.g. "2020-05-07"
\end{itemize}
The observedPeriod is defined as the complete period, which will be analyzed in this project, e.g.,
observedPeriod = 1 + EndData - StartDate.

```{r}
StartDate <- obk$Day[1]
EndDate <- obk$Day[length(obk$Day)] 
observedPeriod = 1 + as.numeric(as.Date(EndDate)- as.Date(StartDate))
```


####  The OBK Simulations/Predictions

```{r}
infectedPerDay <- obk$Infected
infectedPerDayCum <- cumsum(infectedPerDay)
obk$Sick <-  slide_dbl(obk$Infected, ~sum(.x), .before = (amntDaysSickness -1))
# print(data.frame(day = 1:length(infectedPerDay), infected= infectedPerDay, sick = obk$Sick, infectedCum = infectedPerDayCum))
```

```{r}
{plot(obk$Day, obk$InfCum, type="l")
lines(obk$Day, obk$Infected, col = "red")
lines(obk$Day, obk$Sick, col = "blue")
}
```
* TODO: Show results based on Excel percentages

#### How to Perform Simmer Simulations Using Real Data

* We will use the `obk` data again.
* First, we generate `arrivalTimes` for the simulation.
```{r}
str(obk)
set.seed(123)
arrivalTimes <- getArrivalTimes(obk$Infected) 
str(arrivalTimes)
```

* These arrival times are passed to the `babsimHospital` function.

```{r callBabsimHospital, include= TRUE}
conf = babsimToolsConf()

conf$simRepeats = 10
conf$logLevel = 0
conf$seed = 123
y <- babsimHospital(arrivalTimes = arrivalTimes
                   , conf = conf
                   , para = babsimHospitalPara()
 )
```


#### How to Evaluate Results

* To evaluate the simulated results, we extract the real data.
* To prepare real data, we select the real (field) data from the `obk` data set. 
```{r getRealBeds}
 GAData <- data.frame(bed=obk$bed, 
                      intensiveBed=obk$intensiveBed, 
                      intensiveBedVentilation=obk$intensiveBedVentilation,
                      Day = obk$Day)
 GABeds <- getRealBeds(data = GAData,  resource=c("bed", "intensiveBed", "intensiveBedVentilation"))
```


```{r getDailyMaxRes}
res <- getDailyMaxResults(
   conf=conf,
  envs = y,  
  fieldEvents = GABeds)
```

* Now, simulated and real data are available in one single data frame: `res`.

* TODO: Measure Performance
* Error is just the sum of the RMSE values for bed, intensiveBed, and intensiveBedVentilation.

```{r getErrorDef, eval = FALSE}
getError <- function(res){
res1 <- res %>% filter(resource == "bed"  & source == "babsim")
df1 <- unique(data.frame(date = res1$date, x = res1$med))
res2 <- res %>% filter(resource == "bed"  & source == "GA")
df2 <- unique(data.frame(date = res2$date, x= res2$med))
dfBed <- full_join(df1, df2, by=c("date"))
dfBed[is.na(dfBed)] <- 0
rmseBed <- rmse(dfBed[,2], dfBed[,3])

res1 <- res %>% filter(resource == "intensiveBed"  & source == "babsim")
df1 <- unique(data.frame(date = res1$date, x = res1$med))
res2 <- res %>% filter(resource == "intensiveBed"  & source == "GA")
df2 <- unique(data.frame(date = res2$date, x= res2$med))
dfIntensiveBed <- full_join(df1, df2, by=c("date"))
dfIntensiveBed[is.na(dfIntensiveBed)] <- 0
rmseIntensiveBed <- rmse(dfIntensiveBed[,2], dfIntensiveBed[,3])

res1 <- res %>% filter(resource == "intensiveBedVentilation"  & source == "babsim")
df1 <- unique(data.frame(date = res1$date, x = res1$med))
res2 <- res %>% filter(resource == "intensiveBedVentilation"  & source == "GA")
df2 <- unique(data.frame(date = res2$date, x= res2$med))
dfIntensiveBedVentilation <- full_join(df1, df2, by=c("date"))
dfIntensiveBedVentilation[is.na(dfIntensiveBedVentilation)] <- 0
rmseIntensiveBedVentilation <- rmse(dfIntensiveBedVentilation[,2], dfIntensiveBedVentilation[,3])
return(rmseBed + rmseIntensiveBed + rmseIntensiveBedVentilation)
}
```

* We calculate the error as the combined error from the three different bed categories:

```{r getErrorVal, eval = TRUE}
getError(conf=conf, res)
```




#### How to Plot Results From Real Simulation:

```{r}
p <- plotDailyMaxResults(res)
print(p)
```



```{r, eval= FALSE}
ggplotly(p)
```






## Appendix



#### Sick Persons

In contrast to the babsim simmer model, which uses the daily number of infections, 
the standard models calculate the number of \sick persons using a time window of size \amntDaysSickness days.

The number of infected people that require hospitalisation are calculated  as follows:

Not every infected person requires hospitalization.
The number of people, that have to go to the hospital is determined as follows:
Only \infected people from the last \amntDaysSickness days are considered.
These are labeled as \sick.
Note the minus one in the calculation, because the current day counts as one of the \amntDaysSickness days.

```{r ex1InfsCum, eval = TRUE}
amntDaysSickness = 20
Infs <- ex1InfectedDf$Infected
sickPerDay <-  slide_dbl(Infs , ~sum(.x), .before = (amntDaysSickness -1))
head(data.frame(day = 1:length(Infs), infected= Infs, sick = sickPerDay, ex1InfsCum = cumsum(Infs)))
```

```{r ex1Table, echo=FALSE, results='asis', eval= FALSE, include=FALSE}
knitr::kable(head(data.frame(day = 1:length(Infs), infected= Infs, sick = sickPerDay, ex1InfsCum = cumsum(Infs))))
```


```{r plotex1Infs, eval = FALSE}
p <- {plot( cumsum(Infs), type = "l", col="black", xlab = "days", ylab = "infected, sick")
lines( 1:observedPeriod, Infs, col = "blue")
lines( 1:observedPeriod, sickPerDay, col = "red")
legend('topleft', lty = 1, c("infectedCum","infected", "sick"), col = c("black", "blue", "red"))
}
```

* A fixed percentage of the dynamically changing number of sick persons has to go to the hospital, e.g.,
requires a bed, an intensive care bed, or an intensive care bed with ventilation.

For the standard models (OBK, Köln, and WHO) the following percentages, based on the number of \sick persons, were determined:
\factorHospital, \factorIntensive, and \factorVentilation.

The number of patients that need to go to the hospital is based on the infected people from the last \amntDaysSickness days.

```{r, eval = FALSE, include= FALSE}
## Get numbers as estimated in excel table based on fixed coefficients
getDataFromFixedNumbers <- function(factorHospital, factorIntensive, factorVentilation, sickPerDay, source){
    date <- as.POSIXct(seq(as.Date(StartDate), as.Date(EndDate), by="days"))
    return(
        rbind(
            data.frame("resource" = "bed","time" = 1:length(sickPerDay), 
                       "med" = ( sickPerDay * factorHospital), "source" = source,
                       "date" = date),
            data.frame("resource" = "intensiveBed","time" = 1:length(sickPerDay), 
                       "med" = ( sickPerDay * factorIntensive), "source" = source,
                       "date" = date),
            data.frame("resource" = "intensiveBedVentilation","time" = 1:length(sickPerDay), 
                       "med" = ( sickPerDay * factorVentilation), "source" = source,
                       "date" = date)
        )
    )
}
```



# Optimization of the Gamma Shape Parameter:

```{r optimGammaShape, eval= FALSE}
err <- c()
for (i in 1:15){
  trajList <- babsimHospitalTrajectories(GammaShapeParameter = i/10
                   , control = babsimHospitalPara())
  y <- babsimHospital(arrivalTimes = arrivalTimes
                   , simRepeats = 100
                   , control = babsimHospitalPara()
                   , trajList = trajList)
  res <- getDailyMaxResults(
  envs = y,  
  fieldEvents = GABeds)
  err <- c(err,getError(res))
}
```

* Running the previous code chunk takes some time. Therefore we provide the results in the following chunk:

```{r OldresultsOptimGammaShape, eval= TRUE}
errOLD <- c(14.480660, 12.815926, 11.432739, 10.106963,  9.055230,  8.486424,  8.093233,  7.855100,  8.619063,  8.821263,
9.639260, 10.266499, 11.265664, 13.171530, 13.869897)
```

* These are the new results with the bug fixes from ver 1.0.2: 
```{r NewresultsOptimGammaShape, eval= TRUE}
err <- c(14.654427, 13.418457, 11.827037, 10.962369,  9.486851,  8.965150,  8.272998,  7.771752,  8.017022,  8.544688,  9.035368,  9.875065, 10.331149, 11.202670, 12.220776)
```



* Plotting:

```{r plotOptimGammaShape, eval= TRUE, include= TRUE}
plot( (1:15)/10, err, xlab="Gamma Shape Parameter")
```


* Gamma Shape = 0.8 is seems to be a good setting:

```{r GammaShapeParameter08}
para <- babsimHospitalPara()
conf <- babsimToolsConf()
conf$parallel <- FALSE

y2 <- babsimHospital(arrivalTimes = arrivalTimes
                   , conf = conf
                   , para = para
     )
```


```{r}
res2 <- getDailyMaxResults(
   conf=conf,
  envs = y2,  
  fieldEvents = GABeds)
getError(conf=conf, res2)
```
```{r}
p <- plotDailyMaxResults(res2)
print(p)
```





## Summary

\bibliography{../bab-bibfiles/bartzAll.bib}
