---
title: "Hospital Capacity Planning Using Discrete Event Simulation: Analysis" 
author:
  - name: Thomas Bartz-Beielstein
    affiliation: IDEA, TH Köln
    address:
    - Steinmülleralle 1
    - 51643 Gummersbach, Germany
    email:  thomas.bartz-beielstein@th-koeln.de
  - name: Frederik Rehbach
    affiliation: IDEA, TH Köln
    address:
    - Steinmülleralle 1
    - 51643 Gummersbach, Germany
    email:  frederik.rehbach@th-koeln.de
  - name: Olaf Mersmann
    affiliation: IDEA, TH Köln
    address:
    - Steinmülleralle 1
    - 51643 Gummersbach, Germany
    email:  olaf.mersmann@th-koeln.de
  - name: Eva Bartz
    affiliation: Bartz und Bartz GmbH
    address:
    - Goebenstr. 10
    - 51643 Gummersbach, Germany
    email:  eva.bartz@bartzundbartz.de
abstract: >
 Resource planning for hospitals under special consideration of the COVID-19 pandemic.
preamble: |
  % Any extra LaTeX you need in the preamble
  % header-includes:
always_allow_html: true  
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
    fig_caption: yes 
    includes:
          in_header: babsim.hospital.sty
    pdf_document:
vignette: >
  %\VignetteIndexEntry{Hospital Capacity Planning Using Discrete Event Simulation: Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---   


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "%>"
)
```

## Introduction

* Resource and capacity planning for hospitals.
* Paper: A novel modelling technique to predict resource-requirements 
in critical care:  a case study (Lawton, McCooe) see (Lawt19a).


## Packages

```{r installDevtools, eval = FALSE}
## install.packages("devtools")
## devtools::install_github("r-lib/devtools")
```


```{r, installSPOT, eval = FALSE}
url <- "http://owos.gm.fh-koeln.de:8055/bartz/spot.git"
devtools::install_git(url = url)
```


```{r, install babsim.hospital, eval = FALSE}
url <- "http://owos.gm.fh-koeln.de:8055/bartz/babsim.hospital.git"
devtools::install_git(url = url)
```

```{r, cleanStart}

suppressPackageStartupMessages({
library("SPOT")
library("babsim.hospital")
library("simmer")
library("simmer.plot")
library("plotly")
})
```


* Package version of `SPOT` must be larger than `2.0.64`:

```{r loadSPOT, eval = FALSE}
packageVersion("SPOT")
```

# Data used by babsim.hospital

* We combine data from two different sources:
  1. `simData`: simulation data, i.e., input data for the simulation. Here,
  we will use data from the Robert-Koch Institute in Germany.
  2. `fieldData`: real data, i.e., data from the DIVI-Intensivregister. The 
  field data will be used for validating the simulation output. 
  
* Statistically speaking, the `babsim.hospital` simulator models resources 
usage in hospitals, e.g., number of ICU beds ($y$), as a function 
of the number of infected individuals ($x$).

* In addition to the number of infections, information about age and gender will 
be used as simulation input.

* We will take a closer look at these data in the following sections.

# Simulation Data: RKI Data

## Get Data From the RKI Server

* `babsim.hospital` provides a simple function to update the (daily) RKI Data.

```{r, eval = FALSE}
updateRkidataFile("https://www.arcgis.com/sharing/rest/content/items/f10774f1c63e40168479a1feb6c7ca74/data")
```

* The downloaded data will be available as `rkidata`. 

## The Original RKI Data 

* Data from the Robert-Koch Institute is included in the `babsim.hospital` package as `rkidata`:

```{r, getAndShowRkiData}
rkidata <- babsim.hospital::rkidata
str(rkidata)
```

* Copyright Note (quoted from
https://npgeo-corona-npgeo-de.hub.arcgis.com/datasets/dd4580c810204019a7b8eb3e0b329dd6_0):
Die Daten sind die „Fallzahlen in Deutschland“ des Robert Koch-Institut (RKI) 
und stehen unter der Open Data Datenlizenz Deutschland Version 2.0 zur Verfügung. 
Quellenvermerk: Robert Koch-Institut (RKI), dl-de/by-2-0   
Haftungsausschluss: „Die Inhalte, die über die Internetseiten des Robert Koch-Instituts zur 
Verfügung gestellt werden, dienen ausschließlich der allgemeinen Information der Öffentlichkeit, 
vorrangig der Fachöffentlichkeit“.

* RKI Data can be visualized as follows (`region = 0` is Germany, `region = 5` 
North Rhine-Westphalia, `region 5374` Oberbergischer Kreis, etc.):

```{r, visuRki}
ggVisualizeRki(data=babsim.hospital::rkidata, region = 5374)
```

## Preprocessed RKI Data 

* Not all the information from the original `rkidata` data set 
is mandatory for the `babsim.hospital`
simulations. We will only use a subset.

* `babsbim.simulator` provides the function `getRkiData()` that extracts the subset of the
`rkidata` used in our simulation, optimization, and analysis:

```{r, preprocessRki}
rki <- getRkiData(rki = babsim.hospital::rkidata)
str(rki)
```

* As illustrated by the output from above, we use the following data:
  1. `Altersgruppe`: age group (intervals, categories), represented as character string
  2. `Geschlecht`: gender 
  3. `Day`: day of the infection
  4. `IdBundesland`: federal state
  5. `IdLandkreis`: county
  6. `time`: number of days (`0` = start data). It will be used as `arrivalTimes` for the
  `simmer` simulations.
  7. `Age`: integer representation of `Altersgruppe`


# Field Data (Real ICU Beds)

## Get Data From the DIVI Server

* Similar to the `rkidata`, which is available online and can be downloaded from 
the RKI Server, the field data is also available online.

* It can be downloaded from the DIVI Server as follows, where 
`YYY-MM-DD` should be replaced by the current date, e.g, `2020-10-26`:

```{r, getDiviData, eval = FALSE}
updateIcudataFile(
  "https://www.divi.de/joomlatools-files/docman-files/divi-intensivregister-tagesreports-csv/DIVI-Intensivregister_YYYY-MM-DD_12-15.csv") 
```


* The downloaded data will be available in `babsim.hospital` as `icudata`. 

* Note:
  * The `rkidata` download saves one large `rda` file that contains the whole 
  data set (approx 1 MB).
  * In contrast to the download of the `rkidata`, 
the download of the `icudata` data saves only data from one day.

* Warning: If `updateIcudataFile()` is executed several times for the same data, 
there will be duplicates in the resulting file.
* So, please check in advance, if really new data are on the DIVI Server.

* Note: The DIVI data are not open data. The following statement can be
found on the DIVI Webpage: 
" Eine weitere wissenschaftliche Nutzung der Daten ist nur mit Zustimmung der 
DIVI gestattet."

* Data from the DIVI register is included in the `babsim.hospital` package as `icudata`:

```{r, getRkiData}
icudata <- babsim.hospital::icudata
str(icudata)
```

* DIVI Data can be visualized as follows (region = 0 ist Germany, 5 North Rhine-Westphalia, 
5374 Oberbergischer Kreis, etc.)

```{r, viszIcu}
p <- ggVisualizeIcu(region = 5)
```


```{r, viszIcu1}
print(p[[1]])
```

```{r, viszIcu2}
print(p[[2]])
```

## Preprocessing DIVI/ICU Data

* Note: ICU bed without ventilation can be calculated as 
`faelle_covid_aktuell  - faelle_covid_aktuell_beatmet`

* The function `getIcuBeds` converts the 9 dim DIVI ICU data `icudata`
(bundesland,gemeindeschluessel,..., daten_stand) into a data.frame with 
  1. bed
  2. intensiveBedVentilation, and 
  3. Day


```{r}
fieldData <- getIcuBeds(babsim.hospital::icudata)
str(fieldData)
```

* The field data based on DIVI uses two bed categories:
  1. `intensiveBed`: ICU bed without ventilation
  2. `intensiveBedVentilation`: ICU bed with ventilation


# Performing Simulations

* To perform a simulation, the setting must be configured (seed, number of 
repeats, sequential or parallel evaluation, variable names, dates, etc.)

* `region = 5315` represents Cologne.

```{r, selectConfig}
region = 5315
seed = 123
simrepeats = 2
parallel = FALSE
percCores = 0.8
resourceNames =  c("intensiveBed", "intensiveBedVentilation")
resourceEval = c("intensiveBed", "intensiveBedVentilation")
```


* We can specify the field data based on `icudata`  (DIVI) for the simulation as 
follows:

```{r, selectIcuData}
FieldStartDate = "2020-09-01"
# Felddaten (Realdaten, ICU):
icudata <- getRegionIcu(data = icudata,
  region = region)
fieldData <- getIcuBeds(icudata)
fieldData <- fieldData[which(fieldData$Day >= as.Date(FieldStartDate)), ]
rownames(fieldData) <- NULL
icu = TRUE
icuWeights = 1
```


* Next, simulation data (RKI data) can be selected. The simulation data in our example, 
depend on the field data:

```{r, selectRkiData}
SimStartDate = "2020-08-01"
rkidata <- getRegionRki(data = rkidata,
  region = region)
simData <- getRkiData(rkidata)
simData <- simData[which(simData$Day >= as.Date(SimStartDate)), ]
## Auch mit fieldData cutten damit es immer das gleiche Datum ist
simData <-
simData[as.Date(simData$Day) <= max(as.Date(fieldData$Day)),]
## time muss bei 1 starten
simData$time <- simData$time - min(simData$time)
rownames(simData) <- NULL
```

* Finally, we combine all data in one data frame `data`:

```{r, combineSelectedData}
data <- list(simData = simData,
 fieldData = fieldData)

```

* Configuration information is stored in the `conf` list, i.e., 
`conf` refers to the simulation configuration, e.g., 
sequential or parallel evaluation,
number of cores, 
resource names, 
log level, etc.


```{r, getSimConf}
conf <- babsimToolsConf()
conf <- getConfFromData(conf = conf,
                        simData = data$simData,
                        fieldData = data$fieldData)
  conf$parallel = parallel
  conf$simRepeats = simrepeats
  conf$ICU = icu
  conf$ResourceNames = resourceNames
  conf$ResourceEval = resourceEval
  conf$percCores = percCores
  conf$logLevel = 1
  conf$w2 = icuWeights
  set.seed(conf$seed)
```

## Simulation Model Parameters 

* The core of the `babsim.hospital` simulations is based on the `simmer` package. 
* It uses simulation parameters, e.g., arrival times, durations, and 
transition probabilities. 
* These are currently 42 parameters (shown below) that are stored in the list `para`.

```{r, getSimParameter}
para <- babsimHospitalPara()
str(para)
```


# Run simulation

* The `babsim.hospital` simulator requires the specification of
  1. `arrivalTimes`
  2. configuration list `conf`
  3. parameter list `para` 
for the simulation.

* Arrival times were not discussed yet.
* `babsim.hospital` provides the function `getRkiRisk()` that generates arrivals
with associated risks.
* `Risk` is based on age (`Altersgruppe`) and gender (`Geschlect`):

```{r, getRiskAndArrival}
rkiWithRisk <- getRkiRisk(data$simData, para)
head(rkiWithRisk)
```

* To perform simulations, only two parameters are required:
  1. `time`: arrival time 
  2. `Risk`: risk (based on age and gender)
* A data.frame with these two parameters is passed to the main simulation 
function `babsimHospital`.
* Output from the simulation is stored in the variable `envs`.

```{r, runSimmer, echo = FALSE}
arrivalTimes <- data.frame(time = rkiWithRisk$time,
                             risk = rkiWithRisk$Risk)
envs <- babsimHospital(arrivalTimes = arrivalTimes,
                         conf = conf,
                         para = para)
```
## Visualize Output

### Simmer Plots

* First, we illustrate how to generate plots using the `simmer.plot` package.

* In the following graph, the individual lines are all separate replications. 
The smoothing performed is a cumulative average. 

* Besides `intensiveBed` and  `intensiveBedVentilation`, `babsim.hospital` 
also provides information about the number of non-ICU beds. The non-ICU beds are
labeled as `bed`.

* Summarizing, `babsim.hospital` generates output for three bed categories:
 1. `bed`
 2. `intensiveBed`
 3. `intensiveBedVentilation`

* To plot resource usage for three resources side-by-side, we
can proceed as follows:

```{r, plotSimmerThreeResources}
resources <- get_mon_resources(envs)
resources$capacity <- resources$capacity/1e5
plot(resources, metric = "usage", c("bed", "intensiveBed", "intensiveBedVentilation"), items = "server")
```
* Each resource can be plotted separately.

1. The following command generates a plot of non icu beds:

```{r, plotBeds3}
plot(resources, metric = "usage", "bed", items = "server", steps = TRUE)
```

2. The following command generates a plot of icu beds without ventilation:

```{r, plotICU3}
plot(resources, metric = "usage", "intensiveBed", items = "server", steps = TRUE)
```

3. The following command generates a plot of icu beds with ventilation:

```{r, plotVent3}
plot(resources, metric = "usage", "intensiveBedVentilation", items = "server", steps = TRUE)
```

## Evaluate Simulation Results

* `babsim.hospital` provides functions for evaluating the quality of the
simulation results.
* Simulation results depend on the transition probabilities and durations, i.e.,
a vector of more than 30 variables.
* These vectors represent *parameter settings*.
* `babsim.hospital` provides a *default* parameter set, that is based on 
knowledge from domain experts (doctors, members of COVID-19 crises teams, 
mathematicians, and many more).
* We can calculate the error (RMSE) of the default parameter setting, which was used 
in this simulation, as follows:

```{r, getErrsResults}
fieldEvents <- getRealBeds(data = data$fieldData,
                        resource = conf$ResourceNames)
res <- getDailyMaxResults(envs = envs,  fieldEvents = fieldEvents, conf=conf)
resDefault <- getError(res, conf=conf)
```

The error is `r resDefault`.


* Here, we illustrate how `babsim` plots can be generated.

```{r, plotSimAndResults}
p <- plotDailyMaxResults(res)
plot(p)
```

* Using `ggplot` and `plotly`can be used to generate interactive plots.

```{r, interactiveGGplotly}
ggplotly(p) 
```


## Optimization 

* As discussed above, `babsim.hospital` provides a default parameter set, which 
can be used for simulations.
* The function `babsimHospitalPara()` provides a convenient way to access the
default parameter set:
```{r,eval=FALSE}
para <- babsimHospitalPara()
```


* `babsim` provides an interface to optimize the parameter values of 
the simulation model. 
* The following code is just a quick demo.

```{r, optimSpot, eval=FALSE}
### Version: 4.10.26.6
library(babsim.hospital)
library("SPOT")
library("simmer")

runopt(
  expName = "koelnpara20201105PM-",
  rkiwerte = babsim.hospital::rkidata,
  icuwerte = babsim.hospital::icudata,
  region = 5315,
  TrainFieldStartDate = "2020-10-01",
  TrainSimStartDate = "2020-09-01",
  TestFieldStartDate = "2020-10-21",
  TestSimStartDate = "2020-09-21",
  Overlap = 7,
  seed = 101156,
  repeats = 100,
  funEvals = 250,
  funEvalsFactor = 50,
  size = 100,
  simrepeats = 10,
  subset = 100,
  parallel = TRUE,
  percCores = 0.9,
  icu = TRUE,
  icuWeights = c(1,10)
)
```

* Results from the `runopt()` runs are stored as `*.rda` files.
* `babsim.hospital` provides results from the following regions (towns and 
 counties in Germany):
   * `obkpara`: Oberbergischer Kreis
   * `koelnpara`: City of Cologne
   * `nrwpara`: North-Rhine Westphalia
   * `deutschland`: Germany


## Use Optimized Parameters

* Define test data

```{r, cleanStartAndLoadRequiredLibs}

library("babsim.hospital")
library("SPOT")
library("simmer")
library("plotly")
```


```{r, testDefsAndRegion}
# OBK
# region = 5374
# Koeln
# region = 5315
# NRW
region = 5
  rkiwerte = babsim.hospital::rkidata
  icuwerte = babsim.hospital::icudata
  TrainSimStartDate = Sys.Date() - 11*7
  TrainFieldStartDate = Sys.Date() - 8*7
  TestSimStartDate = Sys.Date() - 8*7
  TestFieldStartDate = Sys.Date() - 4*7
  verbosity = 0
  icuWeights = c(1,1)
  testRepeats = 10
  parallel = FALSE
  simrepeats = 10
  icu = TRUE
  resourceNames = c("intensiveBed", "intensiveBedVentilation")
  resourceEval = c("intensiveBed", "intensiveBedVentilation")
  percCores = 0.5
  seed = 1234 
  
  print(c(min(rkiwerte$Refdatum), max(rkiwerte$Refdatum)))    
  print(c(min(icuwerte$daten_stand), max(icuwerte$daten_stand)))    
    
  result.df <- data.frame(x = NULL, y = NULL)
  reslist <- list()
    
   ## General data range adaptation for train and test data:
   ##  Field data = Realdata, ICU, beds:
   regionIcuwerte <- getRegionIcu(data = icuwerte,
                                   region = region)
    fieldData <- getIcuBeds(regionIcuwerte)
    fieldData <-
        fieldData[which(fieldData$Day >= as.Date(TrainFieldStartDate)), ]
    
    ## Sim data = RKI, infections:
    regionRkiwerte <- getRegionRki(data = rkiwerte,
                                   region = region)
    simData <- getRkiData(regionRkiwerte)
    simData <- simData[which(simData$Day >= as.Date(TrainSimStartDate)), ]
  #
  i <- 1
  
```


```{r, getTestPara}
#obk_2020_Nov.21_23.36_V11.0.52R159 <- readRDS("~/workspace/babsim.hospital/babsim.hospital/runs/archive/obk_2020_Nov.21_23.36_V11.0.52R159.rds")
#results <- obk_2020_Nov.21_23.36_V11.0.52R159

## OBK

# SPOT direct:
# load("./runs/results/bartz07/obk-2020-Nov.21-23.36-V11.0.52R214.RData")
# results <- obkpara 


# Koeln:
# SPOT direct:
# load("~/workspace/babsim.hospital/babsim.hospital/runs/results/maans03/koeln-2020-Nov.22-00.13-V11.0.54R62.RData")
# load("~/workspace/babsim.hospital/babsim.hospital/runs/results/koeln_2020_Nov.23_00.38_V11.1.10R57.RData")
# load("~/workspace/babsim.hospital/babsim.hospital/runs/results/koeln_2020_Nov.23_15.35_V11.1.14R12.RData")

# NRW:
# SPOT direct:
# load("~/workspace/babsim.hospital/babsim.hospital/runs/results/maans05/nrw-2020-Nov.21-22.40-V11.0.50R11.RData")
# load("~/workspace/babsim.hospital/babsim.hospital/runs/results/maans04/nrw/nrw_2020_Nov.22_23.23_V11.1.10R10.RData")
#load("~/workspace/babsim.hospital/babsim.hospital/runs/results/nrw_2020_Nov.23_15.46_V11.1.14R1.RData")
# load("~/workspace/babsim.hospital/babsim.hospital/runs/results/nrw_2020_Nov.23_18.10_V11.2.0R1.RData")
# load("~/workspace/babsim.hospital/babsim.hospital/runs/results/nrw_2020_Nov.24_06.50_V11.3.2R1.RData")
load("~/workspace/babsim.hospital/babsim.hospital/runs/results/maans03/nrw/nrwover_2020_Nov.25_07.35_V11.4.2R6.RData")
```


```{r, getTestPara}
results <- result.df
para <- getBestParameter(results)
yColumn <- match('y', colnames(results))
minRow <- which.min(results[, yColumn]) # y column
x <- results[minRow,]
print(x)
```


```{r, eval = FALSE}
results <- result.df
para <- getBestParameter(results)
yColumn <- match('y', colnames(results))
minRow <- which.min(results[, yColumn]) # y column
obkpara <- results[minRow,]
print(as.double(obkpara))
usethis::use_data(obkpara, overwrite = TRUE)
```


```{r, eval = FALSE}
results <- result.df
para <- getBestParameter(results)
yColumn <- match('y', colnames(results))
minRow <- which.min(results[, yColumn]) # y column
nrwkpara <- results[minRow,]
print(as.double(nrwpara))
usethis::use_data(nrwpara, overwrite = TRUE)
##
nrwarchive <- rbind(nrwarchive, result.df)
usethis::use_data(nrwarchive, overwrite = TRUE)
```

```{r, eval = FALSE}
results <- result.df
para <- getBestParameter(results)
yColumn <- match('y', colnames(results))
minRow <- which.min(results[, yColumn]) # y column
koelnpara <- results[minRow,]
print(as.double(koelnpara))
usethis::use_data(koelnpara, overwrite = TRUE)
##
```


```{r, eval = FALSE}
koelnarchive <- rbind(koelnarchive, result.df)
usethis::use_data(koelnarchive, overwrite = TRUE)
```


```{r, prepareTest}
### Entering test:
        print("Starting test evaluation:")
        print("#########################################")
        
        ## Parameters for the test evaluation
        # testPara <- mapXToPara(x)
        
        testPara <- para
        testPara <- checkSimPara(testPara)
        
        ## Generate test data for final evaluation of one optimization run:
        ## testFieldData
        ## testFieldData <-  getIcuBeds(regionIcuwerte)
        testFieldData <-
            fieldData[which(fieldData$Day >= as.Date(TestFieldStartDate)), ]
        rownames(testFieldData) <- NULL
        
        ## testSimData
        ## testSimData <- getRkiData(regionRkiwerte)
        testSimData <- 
            simData[which(simData$Day >= as.Date(TestSimStartDate)), ]
        
        ## v 10.3.6: 
        TestSimStartDate <- min(testSimData$Day)
        
        ## Auch mit fieldData cutten damit es immer das gleiche Datum ist
        ## Synchronize End Date for Sim and Field Data:
        testSimData <-
            testSimData[as.Date(testSimData$Day) <= max(as.Date(testFieldData$Day)),]
        testFieldData <-
            testFieldData[as.Date(testFieldData$Day) <= max(as.Date(testSimData$Day)),]
        
        
        ## v10.3.6: end days synchronized:
        syncedEndTestDate <- min( max(testSimData$Day), max(testFieldData$Day) )
        testSimData <- testSimData[which(testSimData$Day <= syncedEndTestDate), ]
        TestFieldData <- fieldData[which(fieldData$Day <= syncedEndTestDate), ]  
        ###
        
        ## time muss bei 1 starten
        testSimData$time <- testSimData$time - min(testSimData$time)
        rownames(testSimData) <- NULL
        
        ## Combine data
        testData <- list(simData = testSimData,
                         fieldData = testFieldData)
        
        ## Check test data:
        print( paste("testDataSim: ", c(min(testData$simData$Day), max(testData$simData$Day))))
        print( paste("testDataField: ", c(min(testData$fieldData$Day), max(testData$fieldData$Day))))
        SIM_EQ_FIELD_TESTDATA <- 
            (min(as.Date(testData$simData$Day)) == as.Date(TestSimStartDate)) 
        
        if(SIM_EQ_FIELD_TESTDATA == FALSE){print(as.Date(TestSimStartDate))
            stop("babsim.hospital::runoptDirect: Check testData: 
                 TestSimStartDate.")}
        
        
        SIM_EQ_FIELD_TESTDATA <- 
            (min(as.Date(testData$fieldData$Day)) == as.Date(TestFieldStartDate)) 
        if(SIM_EQ_FIELD_TESTDATA == FALSE){
            print(as.Date(TestFieldStartDate))
            stop("babsim.hospital::runoptDirect: Check testData: TestFieldStartDate.")}
        SIM_EQ_FIELD_TESTDATA <- 
            (max(as.Date(testData$simData$Day)) == max(as.Date(testData$fieldData$Day)))
        if(SIM_EQ_FIELD_TESTDATA == FALSE){stop("babsim.hospital::runoptDirect: Check testData:
                                                sim and field data End date differ!")}
        conf <- babsimToolsConf()
        testConf <- getConfFromData(simData = testSimData,
                                    fieldData = testFieldData,
                                    conf = conf)
        testConf$verbosity = verbosity
        testConf$parallel = parallel
        testConf$simRepeats = simrepeats
        testConf$ICU = icu
        testConf$ResourceNames = resourceNames
        testConf$ResourceEval = resourceEval
        testConf$percCores = percCores
        testConf$logLevel = 1
        testConf$w2 = icuWeights
        testConf$seed <- seed + i + 1
        set.seed(testConf$seed)
        
        ## Check test config:
        print( paste("testConfSim: ", c(testConf$simulationDates$StartDate, testConf$simulationDates$EndDate)))
        print( paste("testConfField: ", c(testConf$fieldDates$StartDate, testConf$fieldDates$EndDate)))
        SIM_EQ_FIELD_TESTCONF <- ( (testConf$simulationDates$StartDate != testConf$fieldDates$StartDate) &
                                       (testConf$simulationDates$EndDate != testConf$fieldDates$EndDate))
        SIM_EQ_FIELD_TESTCONF <- 
            (min(as.Date(testConf$simulationDates$StartDate)) == as.Date(TestSimStartDate)) &
            (min(as.Date(testConf$fieldDates$StartDate)) == as.Date(TestFieldStartDate)) &
            (max(as.Date(testConf$simulationDates$EndDate)) == max(as.Date(testConf$fieldDates$EndDate)))
        if(SIM_EQ_FIELD_TESTCONF == FALSE){stop("babsim.hospital::runoptDirect: Check testConf.")}
```


```{r, performTest}
testErr <- 0
        for (j in 1:testRepeats){
            testConf$seed <- seed + i + 1 + j
            set.seed(testConf$seed)
            envs <- modelResultHospital(para = testPara,
                                        conf = testConf,
                                        data = testData)
            testConf$verbosity <- 101
            err <- getError(envs, conf = testConf)
            print(paste( "single test error:", err))
            testErr <- testErr + err
        }
        testErr <- testErr/testRepeats
        print(paste0("testErr: ", testErr))
        print(paste0("babsim.hospital version: ", 
                     utils::packageVersion("babsim.hospital")))
        testResult <- data.frame(y = testErr, x = x)
```


```{r, eval=FALSE}
# obk_meta_2020_Nov.21_23.40_V11.0.52R4 <- readRDS("~/workspace/babsim.hospital/babsim.hospital/runs/archive/obk_meta_2020_Nov.21_23.40_V11.0.52R4.rds")
# results <- obk_meta_2020_Nov.21_23.40_V11.0.52R4
# para <- getBestParameter(results)
res <- modelResultHospital(para = testPara,
                           conf = testConf,
                           data = testData)
resOpt <- getError(res, conf=testConf)
```

* Optimization improves the error
* This improvement can also be visualized.

```{r, eval = FALSE}
p <- plotDailyMaxResults(res)
print(p)
```

* `ggplot` and `plotly`can be used to generate interactive plots.

```{r, showGgplotly, eval = FALSE}
ggplotly(p)
```


## Smooth Parameters

* Smooth a parameter set using another parameter set
* Calculate the average of two parameter sets to smooth out any local anomalies.
*  Mostly useful to smooth out a local (say OBK) parameter set using a global one (say NRW).
* Technically this function calculates `(1-weight) * para + weight * other ` ensuring that the names etc. of  `para` are preserved.
* Parameters:
   * para Parameter set to smooth
   * other Other parameters to average in
   * weight Weight of other parameters
* return Weighted parameter set

```{r smooth, eval = FALSE}
para <- smoothParameter(getBestParameter(babsim.hospital::obkpara), 
                getBestParameter(babsim.hospital::nrwpara) )
```


## Visualize and Analyse Parameter Settings

* `babsim.hospital` includes tools to analyse parameter settings.
* You might recall that parameter settings consist of
   1. transition probabilities, e.g., the probability that an infected
   individual has to go to the hospital. 
   2. durations, e.g., the time span until an infected individual goes to the
   hospital (in days).
* The following plot illustrates the transition probabilities.
* States are as follows:
   1. `infec`: infected
   2. `out`: transfer out, no hospital required
   3. `hosp`: hospital
   4. `normal`: normal station, no ICU
   5. `intens`: ICU (without ventilation)
   6. `vent`: ICU ventilated
   7. `intafter`: intensive aftercare (from ICU with ventilation, on ICU)
   8. `aftercare`: aftercare (from ICU,  on normal station)
   9. `death`: patient dies
  10. `healthy`: recovered
 


* The transition matrix, that stores the probabilities, is shown below:

```{r, showMatrixP}
para <- babsimHospitalPara()
getMatrixP(para = para)
```


```{r, visParaP}
visualizeGraph(para=para, option = "P")
```



* Similar to the probabilites, durations can be visualized:

```{r, showDurationMatrix}
visualizeGraph(para = para, option = "D")
```

* The corresponding matrix is shown below:

```{r, showMatrixDurations}
getMatrixD(para = para)
```


# Extend RKI Data

* `babsim.hospital`can be used to simulate scenarios, i.e., possible 
developments of the pandemic.
* To simulate these scenarios, arrival events must be generated.
* The function `extendRki()` adds new arrival events. 
* To generate new arrivals, three parameters must be specified:
   1. `data`: an already existing data set, i.e., the history
   2. `EndDate`: last day of the simulated data (in the future)
   3. `R0`: base reproduction values (R0) at the first day of the scenario and at the 
   last day of the scenario. A linear interpolation between these two values will 
   be used, e.g., if `R0 = c(1,2)` and ten eleven days are specified, the 
   following R0 values will be used: (1.0, 1.1, 1.2, 1.3, ..., 1.9,2.0).


```{r, extendRkiDemo, eval = TRUE}
data <- getRkiData(babsim.hospital::rkidata)
n <-  as.integer( max(data$Day)-min(data$Day) )
StartDay <- min(data$Day) + round(n*0.9)  
data <- data[which(data$Day >=  StartDay), ]
EndDate <- max(data$Day) + 2
dataExt <- extendRki(data = data, 
                     EndDate = EndDate,
                     R0 = c(0.1, 0.2))
```

* To illustrate the `extendRki()` data extension procedure, a short 
example is shown below:

```{r, eval = TRUE}
visualizeRkiEvents(data = data, region=5315)
```


* The following plot shows the result of the data extension:
 

```{r, eval = TRUE}
visualizeRkiEvents(data = dataExt, region = 5315)
```

# Sensitivity Analysis

## Quick Analysis

* OBK Data

```{r, quickAnalysis1, eval = FALSE}
library("rpart")
library("rpart.plot")
library("babsim.hospital")
library("SPOT")
param <- obkpara_10_4_6
y <- param[,1]
x <- param[,2:dim(param)[2]]
fitTree <- buildTreeModel(x=x,
                 y=y,
                 control = list(xnames = paste0('x', 1:33)))
rpart.plot(fitTree$fit)
```
```{r, eval = FALSE}
getParameterNameList(c(24, 25, 3, 10))
```


```{r, quickAnalysisKoeln, eval = FALSE}
library("rpart")
library("rpart.plot")
param <- koelnpara
y <- param[,1]
x <- param[,2:dim(param)[2]]
fitTree <- buildTreeModel(x=x,
                 y=y,
                 control = list(xnames = paste0('x', 1:33)))
rpart.plot(fitTree$fit)
```


## Sensitivity Analysis

* `babsim.hopital` uses the `R` package SPOT (sequential parameter optimization 
toolbox) to improve parameter settings.
* `SPOT` implements a set of tools for model-based optimization and tuning of 
algorithms (surrogate models, optimizers, DOE).
* `SPOT` can be used for sensitivity analysis, which is in important under 
many aspects, especially:
   * understanding the most important factors (parameters) that influence model
   behavior. For example, it is of great importance for simulation practitioners and
   doctors to discover relevant durations and probabilities. 
   * detecting interactions between parameters, e.g., do durations influence 
   each other?
* Before visualizations are presented, we show the underlying parameter setting.

```{r, eval = TRUE}
res <- res202010262[[2]][[1]]
xBest <- res$xbest
print(xBest)
```
* The corresponding parameter names are:

```{r}
t(getParameterNameList(1:27))
```

* The fitness landscape can be visualized using the function `plotModel`.
* Note, `plotModel` requires two parameter values.
* In the following example, 
`GammaShapeParameter` (x16) and `AmntDaysNormalToHealthy` (x2) were chosen.
* The plot can be interpreted as follows:
   * The model error is reduced, if patients stay longer on the normal station 
   before they leave the hospital (healthy).
   * The effect of the parameter `GammaShapeParameter` is smaller than the effect of 
   the parameter `AmntDaysNormalToHealthy`.


```{r, eval = TRUE}
SPOT::plotModel(res$modelFit, which = c(16,2) ,xlab = c("Modellierungsparameter (Varianz), GammaShapeParameter", "x2: Normalstation zu Genesen (AmntDaysNormalToHealthy)"))
```

* A regression-based parameter screening cn be performed to discover relevant 
(and irrelevant) model parameters:

```{r, eval = TRUE}
fitLm <- SPOT::buildLM(x=res$x,
                 y=res$y,
                 control = list(useStep=TRUE))
summary(fitLm$fit)
```

* Parameter $x_7$  should be considered important.

```{r, eval = TRUE}
getParameterName(7)
```

* This finding is supported by a simple regression tree analysis:

```{r, eval = TRUE}
library("rpart")
library("rpart.plot")
fitTree <- buildTreeModel(x=res$x,
                 y=res$y,
                 control = list(xnames = paste0('x', 1:33)))
rpart.plot(fitTree$fit)
```


## Estimate Risiks

* We consider the model $r(x) = exp(bx)$, where $x$ denotes the input variable, e.g.,
the age, and $b \in R$ is an unknown parameter. 
* We use empirical data to estimate $b$.
* An exponential model with two parameters was chosen to model `risk` as 
a function of `age` and `gender`: 
$ r(x) = a exp(bx)$.

```{r, estimateAB}
age <- c(2,10,25,47,70,90)
    risk <- c(0.01,0.07,0.15,0.65,3,12.64)
    # ab <- getExpCoeff(x=age, y=risk)
    data <- data.frame(x=age, y=risk)
    fit <- nls(y ~ a * exp( b * x), 
               data = data, 
               start = list(a = 1, b = 0),
               control= nls.control(maxiter = 50, tol = 1e-05, minFactor = 1/1024,
                           printEval = FALSE, warnOnly = FALSE)
    )
    print(coef(fit))
```


```{r, plotEstimatedAB}
{plot(age,2*risk)
    # female:
    lines(age, 1* predict(fit, list(x = age)))
    # male:
    lines(age, 2* predict(fit, list(x = age) ), col ="red")}
```

# The RKI Data Classes

## Raw data:  rkidatafull

* The full, unmodified data set, which is downloaded from the RKI web page, 
is stored as `rkidataFull`.
* `rkidataFull` is a large data set:

```{r, dimRkidataFull}
dim(babsim.hospital::rkidataFull)
```


## Class rkidata

* The `rkidata` data set is a subset of the `rkidataFull` data set.
* It contains data from `2020-09-01` until today.
* The `rkidata` subset is used, because the COVID-19 pandemic behavior changed 
over time. The period from September is sometimes referred to as the second wave.

```{r, dimRkidata}
str(babsim.hospital::rkidata)
```

## Class simData

* To convert data from `rkidata` class, the function `getRkiData()` can be used.
* The function generates data that can be used as `simmer` arrival events:
   * one arrival is listed in each row
* Especially, the following information is provided:
   * age group (classification used by RKI)
   * gender 
   * day, the infection was registered (in German: `Meldedatum`)
   * region information (id of the state)
   * region information (id of the county)
   * age as a numerical value (1:1 correspondence with age group)
   
```{r, getSimdataRki}
rkiSimData <- getRkiData(babsim.hospital::rkidata)
str(rkiSimData)
```

### Adding risk information to simData class data

* The function `getRkiRisk( simData, para)` adds a numerical risk value,
which is based on `Age` and `Geschlecht` to  `simData` data.
* The parameters ` x31:RiskFactorA`, `x32: RiskFactorB`, and `x33:RiskMale` 
are used for the risk calculation.

```{r, addRisktoRki}
para <- babsimHospitalPara()
rkiRiskSimData <- getRkiRisk(rkiSimData, para)
str(rkiRiskSimData)
```

## The Class arrivaldata 

* The `babsim.hospital` simulator function `babsimHospital()`, which 
implements a `simmer` class, processes arrival events.
* An element of the `arrivaldata` class can be generrated as follows:

```{r, generateArrivalEvent}
arrivalTimes <- data.frame(time = rkiRiskSimData$time, risk = rkiRiskSimData$Risk)
str(arrivalTimes)
```


# Optimization Details

## Bounds

```{r, eval = FALSE}
getParameterDataFrame()
```


```{r getBounds}
bounds <- getBounds()
print(bounds)
```



# References
\bibliography{../bab-bibfiles/bartzAll.bib}
















